Fast radio burst cosmology: dispersion measure angular power spectrum

### 2015-01-12 writeup/dm_angular_ps.tex
correlation function normalization.
Changes.

% shortcuts
\newcommand{\ud}{\,\mathrm{d}}
\newcommand{\bD}{\boldsymbol D}
\newcommand{\bC}{\boldsymbol C}
\newcommand{\bDelta}{\boldsymbol \Delta}

% for draft comments.
%\newcommand{\red}{\textcolor{red}}
\newcommand{\red}{}
\newcommand{\blue}{\textcolor{blue}}
\newcommand{\green}{\textcolor{green}}

@@

\title{Fast Radio Burst Cosmology}

@@

@@

\subsection{Dispersion measure}

@@

We define the dispersion-measure distance (or dispersion distance),
the equivalent distance required to
produce a given amount of dispersion in an observed radio signal in the
Universe today, as $ D = {{\rm DM}}/{\bar{n}_{e0}}$
We separate the dispersion distance into the contributions from each
term in \ref{e:ne_sep}:
\begin{equation}
\label{e:D_sep}
D = D^{\rm gal} + D^{\bar{n}_e} + D^{\delta_e} + D^{\rm src}.
\end{equation}
For a population of FRBs originating at cosmological distances, ${\rm DM}^{\rm
src}$ is only a single function of $\hat{n}$ which can presumably be well
measured.  In addition, this function could in principal be well characterized
by independent data such as observations of pulsars.  As such, we will
here-after we ignore this term.

@@

\section{2D statistics}
% Need beter section title.

@@

\subsection{The dispersion measure angular power-spectrum}

\subsection{estimating the dispersion measure angular power-spectrum}

\subsection{sensitivity estimates}

\section{3D statistics}

3D clustering of FRBs using DM as distance proxy.

Dispersion measure space distortions.

\subsection{First order DM space distortions}

LOS election perturbations cause clusturing in unclustered FRBs.

\subsection{Second order DM space distortions}

LOS election perturbations cause non-stationarity/non-gaussianity in clustered FRBs.

@@

\section{Sensitivity}

@@

\section{Simplified sensitivity calculation}

>>>>

% shortcuts
\newcommand{\ud}{\,\mathrm{d}}
\newcommand{\bD}{\boldsymbol D}
\newcommand{\bC}{\boldsymbol C}
\newcommand{\bDelta}{\boldsymbol \Delta}
\newcommand{\Dgal}{D^{\rm gal}}
\newcommand{\Dne}{D^{\bar{n}_e}}
\newcommand{\Dde}{D^{\delta_e}}
\newcommand{\Dsrc}{D^{\rm src}}
\newcommand{\del}{\delta\!}
\newcommand{\calP}{{\cal P}}

% for draft comments.
\newcommand{\red}{\textcolor{red}}
\newcommand{\blue}{\textcolor{blue}}
\newcommand{\green}{\textcolor{green}}

@@

\title{Fast radio burst cosmology: dispersion measure angular power spectrum}

@@

In this paper we consider only the simplest statistic that captures some of
the cosmological information present in an FRB survey.  This is the dispersion
measure angular power-spectrum, which is conceptually in vary close analogy to
the galaxy weak-lensing shear angular power spectrum. We start with a
population of FRB events (the analogue of lensed galaxies) whose redshifts are
unknown.  In the first instance we ignore the intrinsic clustering of these
background sources.  The principal observable is the dispersion measure
(shear) which is a line-of-sight integral of the intervening large-scale
structure.  This is contaminated by a noise term (analogue of intrinsic
ellipticity) which is the dispersion
measure originating from the host halo as well as the dispersion measure
originating from the spacial mean electron density coupled with the unknown
redshift.  In both cases there is a systematic effect originating from the
intrinsic clustering of source objects, which is that pairs of sources with
small angular separation are likely associated with the same large-scale
structure and thus have similar redshifts.  This induces correlations in the
noise tern associated with the unknown distance through the mean electron
density. The analogue in galaxy lensing is intrinsic alignments, however while
intrinsic alignments are a non-linear effect, the redshift clustering term occurs
at linear order.

@@

@@

We define the dispersion-measure distance (or dispersion distance),
the equivalent distance required to
produce a given amount of dispersion in an observed radio signal in the
Universe today, as $ D = {{\rm DM}}/{\bar{n}_{e0}}$
We separate the dispersion distance into the contributions from each
term in \ref{e:ne_sep}:
\begin{equation}
\label{e:D_sep}
D = \Dgal + \Dne + \Dde + \Dsrc
\end{equation}
For a population of FRBs originating at cosmological distances, ${\rm DM}^{\rm
src}$ is only a single function of $\hat{n}$ which can presumably be well
measured.  In addition, this function could in principal be well characterized
by independent data such as observations of pulsars.  As such, we will
here-after we ignore this term.

In addition, we define the mean subtracted values of the of the variouse dispersion
measure quantites, $\del D = D - \langle D \rangle$. Here the angular brackets represent
the ensemble average.

@@

\red{Introduce the redshift distribution function, assumption that FRB's trace
LSS, bias, etc.}

\red{Convensions: we use subscripts $i$, $j$, etc.~on coordinate-like
    quantities to label observed FRB events. Thus $\vec x_i$ and $\hat{n}_i$
    are respectively the 3D coordinates of and unit vector pointing to event $i$.
    This should not be confused with superscripts, denoting the
component of a vector, such that $x^1$ is the first componenent of vector $\vec
x$.  }

\red{$\calP$ is a probability, while $p$ is a PDF. $\calP(x < X < x + \ud x) =
p(x) \ud x$ always.}

\section{Dispersion measure angular correlations}

The principle observable which we would like to consider is the correlations
between the dispersion measures of observed fast radio bursts as a function of
their angular separation.  For observed FRB events labeled by the indices $i$
and $j$, and assuming $i \neq j$, we define the dispersion measure angular
correlation function as
\begin{equation}
    w(\hat{n}_i\cdot\hat{n}_j) \equiv \langle \del D_i \del D_j \rangle.
\end{equation}
A move convienient quantity to work with is the corresponding angular
power spectrum
\begin{equation}
w(\hat{n}_i \cdot\hat{n}_j) = \sum_l \frac{2l+1}{4 \pi} C_l
P_l(\hat{n}_i\cdot\hat{n}_j).
\end{equation}

We have identified two effects that dominate the angular
correlations and which depend on the large-scale structure of the Universe and
thus cosmology.  The first effect is that the signals from two FRB events may
propagate through shared structures along the line of sight.  Thus correlations
in the election overdensities along the line of sight will correlate the 
$\Dde$.  The second effect is that due to the intrisic clustering of source
events, a pair of FRBs that are nearby in the angular coordinates are more
likely to also be nearby in the line of sight coordinate, thus causing
correations in $\Dne$.  In the remainder of this section we will calculate the
contributions from each of these effects.

\subsection{Correlations in $\Dde$}

\begin{equation}
C_l^{zz'}=\frac{2}{\pi} \int_0^{\infty}k^2 dk P_{ee}(k) \int_0^{\eta^z}\frac{d\eta}{a(\eta)} \int_0^{\eta^{z'}}\frac{d\eta'}{a(\eta')}  j_l(k\eta) j_l(k\eta')
\end{equation}
\begin{equation}
C_l^{\Dde \Dde} = \int d\eta^z \int d\eta^{z'} {\cal P}(\eta^z){\cal P}(\eta^{z'}) C_l^{z z'}
\end{equation}

\subsection{Correlations in $\Dne$}

The joint probablility of observing a pair of FRB events at 3D coordinates
$\vec x_i$ and $\vec x_j$ in infintessimal volumes $\ud^3\vec x_i$ and
$\ud^3\vec x_j$ is porportional to
\begin{equation}
    \calP(\vec x_i, \vec x_j) = \calP(\eta_i) \calP(\eta_j) \left[ 1 + b(z_i)
    b(z_j) \xi(\vec x_i - \vec x_j) \right] \frac{\ud^3\vec x_i}{\eta_i^2}
    \frac{\ud^3\vec x_j}{\eta_j^2}.
\end{equation}
The factors of $1/\eta^2$ are nessisary to achive the proper volume weighting,
a fact that can be seen by taking the limit of $\xi \to 0$ and integrating of
the angular coordinates.

An observed event has known angular coordinates, and we are interested in
correlations in $\Dne$ which is directly related to the line of sight
coordinate.  Fixing the angular coordinates:
\begin{equation}
    \calP(\eta_i, \eta_j | \hat{n}_i, \hat{n}_i) = N(\hat{n}_i, \hat{n}_i)
    \calP(\eta_i) \calP(\eta_j) \left[ 1 + b(z_i)
    b(z_j) \xi(\vec x_i - \vec x_j) \right] \ud \eta_i \ud \eta_j,
\end{equation}
where $N(\hat{n}_i, \hat{n}_i)$ is a normalizing factor which ensures the
expression integrates to unity over $\eta_i$ and $\eta_j$. \red{How to deal
with the $1/\eta^2$ factors?.}
Transforming to $\Dne$ as our distance measure, we obtain
\begin{equation}
\calP(\Dne_i, \Dne_j | \hat{n}_i, \hat{n}_i) = \calP(\eta_i, \eta_j |
\hat{n}_i, \hat{n}_i) \frac{\ud }{}
\end{equation}

\subsection{2D statistics}
% Need beter section title.

@@

\subsubsection{The dispersion measure angular power-spectrum}

\subsubsection{estimating the dispersion measure angular power-spectrum}

\subsubsection{sensitivity estimates}

\subsection{Random notes about correlation function normalization}

\begin{equation}
\langle D_i \rangle =  \langle D^{\rm src}_i \rangle
    + \langle D^{\bar{n}_e}_i \rangle
\end{equation}

If $i \neq j$:
\begin{align}
\langle D_i D_j \rangle - \langle D_i \rangle \langle D_j \rangle
    =& \langle (D^{\delta_e}_i + D^{\rm src}_i +  D^{\bar{n}_e}_i)
              (D^{\delta_e}_j + D^{\rm src}_j +  D^{\bar{n}_e}_j) \rangle
              - \langle D_i \rangle \langle D_j \rangle
    \\
    =& \langle [
        D^{\delta_e}_i
        + (D^{\rm src}_i - \langle D^{\rm src}_i \rangle)
        + \langle D^{\rm src}_i \rangle
        +  (D^{\bar{n}_e}_i -  \langle D^{\bar{n}_e}_i \rangle)
        + \langle D^{\bar{n}_e}_i \rangle
    ]
    \nonumber \\
    & \qquad [
        D^{\delta_e}_j
       + (D^{\rm src}_j - \langle D^{\rm src}_j \rangle)
        + \langle D^{\rm src}_j \rangle
        +  (D^{\bar{n}_e}_j -  \langle D^{\bar{n}_e}_j \rangle)
        + \langle D^{\bar{n}_e}_j \rangle
    ] \rangle
    \nonumber \\
    & \qquad - \langle D_i \rangle \langle D_j \rangle
    \\
    =& \langle D^{\delta_e}_i D^{\delta_e}_j \rangle
        + \langle (D^{\rm src}_i - \langle D^{\rm src}_i \rangle)
                 (D^{\rm src}_j - \langle D^{\rm src}_j \rangle) \rangle
        + \langle (D^{\bar{n}_e}_i -  \langle D^{\bar{n}_e}_i \rangle)
                  (D^{\bar{n}_e}_j -  \langle D^{\bar{n}_e}_j \rangle) \rangle
    \nonumber \\
    & \qquad
        + \langle D^{\rm src}_i \rangle \langle D^{\rm src}_j \rangle
        + \langle D^{\bar{n}_e}_i \rangle \langle D^{\bar{n}_e}_j \rangle
        + \langle D^{\rm src}_i \rangle \langle D^{\bar{n}_e}_j \rangle
        + \langle D^{\bar{n}_e}_i \rangle \langle D^{\rm src}_j \rangle
    \nonumber \\
    & \qquad - \langle D_i \rangle \langle D_j \rangle
    \\
    =& \langle D^{\delta_e}_i D^{\delta_e}_j \rangle
        + \langle (D^{\rm src}_i - \langle D^{\rm src}_i \rangle)
                 (D^{\rm src}_j - \langle D^{\rm src}_j \rangle) \rangle
        + \langle (D^{\bar{n}_e}_i -  \langle D^{\bar{n}_e}_i \rangle)
                  (D^{\bar{n}_e}_j -  \langle D^{\bar{n}_e}_j \rangle) \rangle
\end{align}
In the last equation, the first term is the term we've been dealing with, the
second term is hopefully zero, and the third term is what Ue-Li is worried
about, which we still have to calculate. I don't see anything unexpeced.

@@

\subsection{Sensitivity}

@@

\subsection{Simplified sensitivity calculation}
